---
title: "Supporting Portal Predictions"
output: rmarkdown::html_vignette
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{codebase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(portalcasting)
vers <- packageVersion("portalcasting")
today <- Sys.Date()
```

This vignette outlines the codebase for the **portalcasting** package
(v`r vers`), which underlies the automated iterative forecasting within the
[Portal Predictions repo](https://github.com/weecology/portalPredictions).

## Installation

To obtain the most recent version of **portalcasting**, install the most recent 
version from GitHub:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("weecology/portalcasting")
```


## Function pipeline

The package organizes the main actions into a set of hierarchical functions 
`setup_dir()` (which creates and fills the directory), `portalcast()` (which
conducts forecasting or hindcasting as requested), and `cleanup_dir()`
(which removed temporary folders). The functions are designed to operate 
out-of-the-box with no need to alter arguments for operation, in particular 
within the context of the Portal Predictions repo. However, the functions
are flexible, as detailed in [Options List](#options_list).

* `setup_dir()`
  * `create_dir()`
    * `create_main_dir()`
    * `create_sub_dirs()`
      * `create_sub_dir()`
  * `fill_dir()`
    * `fill_PortalData()`
	  * `download_observations()`
    * `fill_data()`
      * `transfer_hist_covariate_forecasts()`
      * `transfer_trapping_table()`
	  * `prep_moons()`
	  * `prep_rodents()`
	  * `prep_covariates()`
	  * `prep_metadata()`
	* `fill predictions()`
	* `fill_models()`
	  * `write_model()`

* `portalcast()`
  * `clear_tmp()`
  * `verify_models()`
  * `prep_data()`
  * `casts()`
    * `cast()`
      * `check_to_skip()`
      * `cast_models()`
      * `combine_forecasts()`	  
      * `add_ensemble()`
      * `clear_tmp()`
	  * `step_casts()`
        * `step_hind_forward()`
        * `update_data()`
          * `prep_moons()`
          * `update_rodents()`
          * `update_covariates()`
          * `prep_metadata()`
        * `cast()`	
          * `check_to_skip()`
          * `cast_models()`
          * `combine_forecasts()`	  
          * `add_ensemble()`
          * `clear_tmp()`

* `cleanup_dir()`

## Directory Tree

The package uses a simple directory tree to organize the project. There are 
three levels to the hierarchy:

* `base`: existing folder where the project folder will be housed
  * `main`: project folder encompassing all subfolders
    * `subs`: specific sub-folders for the project
 
The tree is housed in a simple list, created by the `dirtree()` function, which 
allows for the renaming of the folders at all levels, although it is only 
advisable to alter the `base` or `main` levels, as many functions require 
specifically-named subdirectories. 

The full-path components of the tree can be generated from the simple 
list using `main_path()`, `sub_path()`, and `sub_paths()`. In addition,
`model_path()` can be used to find a specific model's full path
and `file_path()` can be used to find the full path of any file within
the directory.

## Options List <a name="options_list"></a>

The options list is a hierarchical list containing a suite of inter-linked
options controlling the structure, content, and usage of the directory.

* `options_all`: created by `all_options()` 
  * `options_dir`: created by `dir_options()`
  * `options_PortalData`: created by `PortalData_options()`
  * `options_data`: created by `data_options()`
    * `moons`: created by `moons_options()`
    * `rodents`: created by `rodents_options()`
    * `covariates`: created by `covariates_options()`
    * `metadata`: created by `metadata_options()`
    * additionally includes `cast_type`, `quiet`, and `tree` options
  * `options_predictions`: created by `predictions_options()`
  * `options_models`: created by `models_options()`
  * `options_casts`: created by `casts_options()` 

In addition, model-specific scripts (housed within `models/`) are generated 
using options lists, based on the `model_options()` template (not to be 
confused with `models_options()`). 