---
title: "Getting Started"
subtitle: "Set Up and Use a Portal Predictions Directory"
output: rmarkdown::html_vignette
author: "Juniper L. Simonis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r, include=FALSE}
  library(portalcasting)
  vers <- packageVersion("portalcasting")
  today <- Sys.Date()

  models_controls <- prefab_models_controls( )
  models_names    <- unlist(mapply(getElement, prefab_models_controls(), "metadata")["print_name", ])

  nmodels        <- length(models_names)
  nmodels_text   <- as.character(english::english(nmodels))


```


This vignette shows how to set up and run your own [Portal Predictions](https://github.com/weecology/portal-forecasts) directory.

## Installation

First things first, make sure you have the current version of **portalcasting** installed from GitHub:

```r
install.packages("remotes")
remotes::install_github("weecology/portalcasting")
library(portalcasting)
```

## Create a Portal Predictions directory

The `setup_dir()` function creates and populates a standard Portal Predictions directory that includes `resources`, `data`, `models`, `fits`, and `forecasts` subdirectories. 
By default, `setup_dir()` downloads the most up-to-date version of the [Portal Data](https://doi.org/10.5281/zenodo.1215988) archive from GitHub and the NMME downscaled weather forecasts into the `resources` subdirectory; unpacks relevant components; prepares the newmoons, rodents, covariates, covariate forecast, and metadata data files (and places them in the `data` subdirectory); and populates the `models` subdirectory with scripts for the `r nmodels_text` existing models (`r models_names`).

Specialized versions of `setup_dir` are tailored for local model exploration, development, and  testing ("`setup_sandbox`") and for use in the real-deal pipeline ("`setup_production`"). The settings are fairly similar, although `setup_sandbox` has extra rigid argument checking turned off and extra verbose messaging turned on. In addition, it does not download the repository archive resource (only the Portal Data and NMME resources). `setup_production` provides a robust starting point for a user interested in seeing the range of what the package can do, with specific error checking and verbose messaging. `setup_production` does download the entire [Portal Predictions](https://doi.org/10.5281/zenodo.833438) archive from GitHub. Note that downloading the full directory does take a few minutes.

There are many arguments available for the user to tailor the setup of the directory, which can be found on the help pages (`?setup_dir` and `?directory_settings`). Perhaps the most important argument is `main` which allows the user to point the directory to any particular location. The default is `main = "."`, which is basically "the present directory". A common implementation would be to create a directory in the home directory of a computer (indicated by `"~"`) and within a named folder, say "portalcast_directory", which would be done with `main = "~/portalcast_directory"`, or by setting `main <- "~/portalcast_directory"` and then using `main = main` throughout the code:

```{r, eval=FALSE}
main <- "~/portalcast_directory"
setup_dir(main = main)
```

The `models` and `datasets` arguments to `setup_<>` functions allow for control over which models and rodent data sets are included in the directory, and new models and datasets can be added as long as they are given controls in the `new_dataset_controls` or `new_model_controls` arguments. 
Many additional arguments for altering directory set up are available indirectly via the `settings` argument and wrapped into the `directory_settings` functions.

## Run a forecast

The `portalcast()` function controls the running of potentially multiple models across various species in different data sets for a given time setup (defined in `directory_settings` and `time_settings` specifically). This can involve what we have classically called "forecasts" (from the present time step) and "hindcasts" (from a time step before the current one), although the data must be set accordingly (using `time_settings`) for each time point prior to invoking `portalcast()`.

Presently, the preloaded model set includes `r nmodels_text` models: `r models_names`. The jags and GARCH models all take a bit longer to run, so for the purposes of illustration, we will only run ESSS, AutoArima, and NaiveArima models, which we indicate through input to the `models` argument: 

```{r, eval=FALSE}
portalcast(main = main, models = c("ESSS", "AutoArima", "NaiveArima"))
```

If the user does not specify the models, all of the prefab models are run. Note that we need to point `portalcast` to the directory of interest via the `main` argument. This allows us to go between different directories from the same R session with relative ease, but it does mean that `main` is a key argument in nearly all functions. Indeed, `main` is the *only* argument for many **portalcasting** functions.

## Plot the results

Presently two plotting types are available for visualizing the data and model results: time series and point-in-time plots for forecasts. The functions for both of these figure types point directly to the forecast metadata file that allows for flexible selection of which specific model, data set, and end moon (forecast origin) to use, as well as selection via specific identifiers (when multiple versions of a model are run). 

Time series plots are constructed using `plot_forecast_ts`:

```{r, eval=FALSE}
plot_forecast_ts(main = main, dataset = "controls")
```

Point-in-time prediction plots are constructed using `plot_forecast_point`, and default to the next step ahead in time:

```{r, eval=FALSE}
plot_forecast_point(main = main, dataset = "controls")
```




## Reading in data to the R console from the directory

A series of `read_<name>` functions are available for simple loading of the data sets into R from the directory. A generalized `read_data` function includes an  argument for which data set to load ("rodents" [and then which specific data set], "covariates", "climate_forecasts", "newmoons", or "metadata"), and each of those data sets also has a specific function, such as `read_moons`. `read_forecasts_metadata` has a function itself, but is not called via `read_data` currently.

```{r, eval=FALSE}
read_data(main = main, data_name = "rodents")
read_data(main = main, data_name = "rodents_dataset", dataset = "all")
read_data(main = main, data_name = "rodents_dataset", dataset = "controls")
read_data(main = main, data_name = "covariates")
read_data(main = main, data_name = "climate_forecasts")
read_data(main = main, data_name = "newmoons")
read_data(main = main, data_name = "metadata")

read_rodents(main = main)
read_rodents_dataset(main = main)
read_covariates(main = main)
read_climate_forecasts(main = main)
read_newmoons(main = main)
read_metadata(main = main)
read_forecasts_metadata(main = main)
```

## Reading in forecast output to the R console from the directory

Presently, four functions are available for interfacing with saved forecast output. 

`select_forecasts` provides a simple interface to the forecast metadata file with quick filtering:

```{r, eval=FALSE}
select_forecasts(main = main, models = "AutoArima")
```

And `read_forecast_tab`, `read_model_forecast`, and `read_forecasts_metadata` each read in the associated output from a given forecast, as indicated by its forecast_id, which is displayed in the output from `select_forecasts`:

```{r, eval=FALSE}
read_forecast_table(main = main)
read_model_forecast(main = main)
read_forecast_metadata(main = main)
```

