---
title: "Set Up and Use a Portal Predictions Directory"
output: rmarkdown::html_vignette
author: "Juniper L. Simonis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{howto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(portalcasting)
vers <- packageVersion("portalcasting")
today <- Sys.Date()
```

This vignette shows how to set up and run your own [Portal 
Predictions](https://github.com/weecology/portalPredictions) directory.

### Create a Portal Predictions directory

The `setup_dir()` function creates and populates a standard Portal Predictions
directory within the present location (by default) that includes `data`,
`models`, `PortalData`, `predictions` and `tmp` subdirectories. By default,
`setup_dir()` downloads the most up-to-date version of the 
[Portal Data from Zenodo](https://zenodo.org/record/2541170)
into the `PortalData` subdirectory, prepares the moons, rodents, covariates,
and metadata data files (and places them in the `data` subdirectory), copies 
the historic covariate forecast data file available in the package 
(`/inst/extdata/covariate_forecasts.csv`) into the `data` subdirectory, 
downloads the most recent set of model forecasts into the `predictions` 
subdirectory and populates the `models` subdirectory with scripts for the four
existing models ("AutoArima", "ESSS", "nbGARCH", and "pevGARCH").

```{r, eval=FALSE}
setup_dir()
```

### Run a "-cast" (forecast or hindcast)

The `portalcast()` function controls the running of potentially multiple 
models for either a forecast or a hindcast. It will prepare the data 
(rodents, covariates, newmoons, metadata) as needed, verify that the requested
models are available to run, run the models, compile the output, and
generate an ensemble (if desired). Presently, the preloaded model set includes
[four models](https://weecology.github.io/portalcasting/articles/models.html):
ESSS, AutoArima, nbGARCH, and pevGARCH. 

```{r, eval=FALSE}
portalcast()
```

### Tidy up the temporary files

The `cleanup_dir()` function simply removes the temporary files and folders
from the directory.

```{r, eval=FALSE}
cleanup_dir()
```

### Options

All required inputs to `setup_dir()`, `portalcast()`, and `cleanup_dir()`
are encompassed within the singular argument `options_all`, which is a 
hierarchical list of control arguments, defined through the options-setting 
function `all_options()` and the default values are appropriate for basic 
forecast usage. There are many options available for the user to and a full 
list can be found by running `?all_options`, to return the help file for 
the functions that generate the hierarchical options list. 

### Loading the model output into an R session

Once the models have finished running, you can read the results back into R
using `read_cast()`, which loads the most recent model forecast or hindcast,
or `read_casts()`, which loads all forecasts or hindcasts:

```{r, eval=FALSE}
cast <- read_cast()
casts <- read_casts()
```

Both functions have simple controls for forecasts or hindcasts and which files
to load based on their date.

### Selecting a subset of model output

The model output that has been loaded can be subset using `select_casts()`,
which has arguments to allow filtering based on `species`, `level`,
`model`, and `newmoonnumber`, with the default being to return all values:

```{r, eval=FALSE}
select_casts(casts, species = "total", model = "Ensemble", level = "All")
```

### Visualizing the model output

Basic plotting of a single-species forecast can be executed using 
`plot_cast_ts()`, which defaults to plotting the total abundance for the 
control plots under the ensemble model, but has arguments to allow full 
flexibility.

```{r, eval=FALSE}
plot_cast_ts()
```

Similarly, the point-in-time forecasts for multiple species can be compared
using `plot_cast_point`, which defaults to all species and the total 
abundance one step ahead for the control plots under the ensemble model,
but has arguments to allow full flexibility.

```{r, eval=FALSE}
plot_cast_point()
```

It is possible to compare the point-in-time forecasts with observed data
by setting `with_census = TRUE`: 

```{r, eval=FALSE}
plot_cast_point(with_census = TRUE)
```

### Visualizing model evaluation

The models are evaluated using a variety of error metrics, which can be plotted
using `plot_err_lead_spp_mods()` and `plot_cov_RMSE_mod_spp()`. 

`plot_err_lead_spp_mods()` shows the raw forecast error as a function of lead time
across the combination of species and models:

```{r, eval=FALSE}
plot_err_lead_spp_mods()
```

The arguments are set up by default to plot a set of recent forecasts for a set 
of common species in the control plots, but they allow for flexibility

Similarly, `plot_cov_RMSE_mod_spp()` displays the coverage and root mean squared 
error (RMSE) of each model across species:

```{r, eval=FALSE}
plot_cov_RMSE_mod_spp()
```

By default, the function produces a figure using all available hindcasts for a 
set of common species in the control plots, but the arguments allow for flexibility.
